### 2025-04-04 latex/HP2_comm.tex
new files

\begin{versinnote}
\tl{atha prāṇāyāmaḥ--\\+}
\tl{athāsane dṛḍhe yogī vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
\end{versinnote}

@@

\begin{versinnote}
\tl{atha nāḍīśuddhiḥ--\\+}
\tl{tathā coktaṃ yogacandrikāyām--\\+}
\tl{athāsane dṛḍhībhūte vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp02_001]
%\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc. (not worth commenting on as alpha has dṛḍhe).
The \emph{Jyotsnā} (2.1) has the plural \emph{prāṇāyāmān}, which is supported by some manuscripts of the \textbeta, \textgamma\ and \textepsilon\ groups. The plural is possible here as it could refer to the different techniques of retention (\emph{kumbhaka}) taught later in the chapter. This is how Brahmānanda understands it (\emph{prāṇāyāmān vakṣyamāṇān}). The variation between singular and plural recurs through this chapter, and we have followed the \textalpha\ readings, which make good sense. In this case, since the verse is introducing the topic of breath control, the more general sense of the singular is appropriate.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind moves, everything moves and, when it is still, everything is still, so the yogi attains motionlessness through restraining the breath.
\end{translation}

@@

\begin{philcomm}[hp02_002]
%The \textalpha\ manuscripts, as well as those of various other important groups, have \emph{sarvaṃ} instead of \emph{cittaṃ} in the first verse quarter. The stemma aside, the former makes better sense as a general statement (i.e., when the wind moves, everything moves, etc.). The reading \emph{citta} only seems to make sense if one understands \emph{sthāṇutvam} as referring to \emph{samādhi} (i.e., `motionlessness [of the mind]'). As far as we know, such a meaning of \emph{sthāṇutvam} is not attested elsewhere. Nonetheless, Brahmānanda glosses \emph{sthāṇutvam} as \emph{kāṣṭhavat} (`like a piece of wood'), implying that it refers to a yogi in \emph{samādhi} as this simile is found elsewhere in contexts of meditative absorption (e.g., \emph{Amanaska} 1.27).
%(JM: but we have sarvam now and cittam is clearly wrong; I think this note is unnecessary — sthāṇu can mean a piece of wood or a post and is a byword for motionlessness (but never means samādhi), like kāṣṭha, and we don't want to put people off reading the notes by including too many unedifying ones.) 
% JB: Since we often read against the stemma, I dont think it's very convincing to simply say we are reading with the stemma in cases where there are two well-attested plausible readings. In this particular case, cittaṃ is a plausible reading that's well attested among the collated mss., so I think we should firstly bring it to the reader's attention and then explain why we think the adopted reading is better on semantic grounds. Perhaps, my attempts to do the latter (by tidying up a previous note) can be improved?
%JM: should we mention that sthāṇu can mean Śiva?
In the second verse quarter, most of the \textalpha, \textbeta\ and \textgamma\ manuscripts have \emph{dṛḍhabandhanaṃ} instead of the reading we have adopted, \emph{niścalaṃ tathā}, which is supported by some \textbeta, \textdelta\ and \texteta\ manuscripts, and also the source text (i.e. the \emph{Vivekamārtaṇḍa}) and testimonia. The reading \emph{dṛḍhabandhanaṃ} appears to be a dittographical error and does not make sense when read with either \emph{sarvaṃ} (\textalpha) or \emph{cittaṃ} (\textgamma).

As noted by Brahmānanda, in addition to its primary meaning of immobility, \emph{sthāṇutvam} can also mean the state of being Śiva, for whom Sthāṇu is another name.

@@

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate |\\+}
\tl{maraṇaṃ tasya niḥkrāntau tato vāyuṃ nirodhayet ||\\+}
\tl{\var{72a sthito ] sthiro A 72b jīvitam ] VAGBGPT; jīvanam GLGPk 72c niṣkrāntau ] VTA; niḥkrāṃtaṃ GBGP, niṣkrāntis GLGPk 72d vāyuṃ nirodhayet ] T; vāyunibandhanāt V, vāyuṃ nibandhayet ATvl, vāyuṃ nirundhayet G}\\!}
\end{versinnote}

@@

\begin{versinnote}
\tl{vyāpārād yasya ceṣṭante śārīrāḥ pañca vāyavaḥ ||\\+}
\tl{prāṇāpānād ayas te tu bhinnā vṛtter na vastutaḥ |\\+}
\tl{vṛttiṃ leśān nigadato bharadvāja nibodha me ||\\+}
\tl{vṛttiḥ praṇayanaṃ nāma yat taj jīvanam ucyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_003]

\emph{Haṭharatnāvalī} 3.80

@@

We believe that \emph{nirundhayet} was probably the original reading and that it was changed to the more correct form of \emph{nirodhayet} and, in some cases, confused with the final verse quarter of the previous verse and changed to \emph{nibandhanāt} (\textalpha) or \emph{nibandhayet} (\textbeta).%JM: necessary note?
%JB nirundhayet is not supported by alpha. 
% JB I've commmented on jīvitaṃ/jīvanam because it looks like its unsupported by alpha (alpha one has jītaviṃ). I now see that J5 has jīvitaṃ, so we could eliminate the note if we add this to the collation. But a note might still be helpful here because G4 is damaged at this place, so jīvitaṃ is not so well attested and jīvanaṃ is possible. 
\end{philcomm}

@@

%\begin{philcomm}[hp02_004]
%The reading \emph{kāyasiddhi} in the final verse quarter is supported by manuscripts of the \textalpha, \textbeta, \textepsilon\ and \texteta\ groups whereas other manuscripts have \emph{kāryasiddhi}. In the context of physical yoga, \emph{kāyasiddhi} makes better sense, as \emph{kāryasiddhi} can refer more generally to accomplishing anything.
%JM: necessary for the reader? Goes with the stemma, makes better sense, variant not particularly remarkable. But maybe I have become too parsimony-partial.
% JB. In this case I agree!
%\end{philcomm}

@@

\begin{philcomm}[hp02_005]
%We have understood the term \emph{nāḍīcakra} as the network of channels that originate from the bulb in the lower abdomen. This interpretation is consistent with \emph{Sārdhatriśatikālottara} 10.1–3:%JM there are many many possible refs for this, do we need one? And some systems have them coming from the navel, and some from the heart (Shaman's new article).

@@

\begin{philcomm}[hp02_006]
%There is a problem in the third verse quarter. Among the divergent readings, the delta manuscripts convey the idea that the impurities are in the central channel  (\emph{suṣumṇāntarasthā}), which makes a defective \emph{ra-vipulā} (there is no caesura after the fourth syllable). This is similar in sense to V2(siglum), which has the reading \emph{suṣumṇā\-madhya\-sthā}. This meaning was accepted by Brahmānanda, who adopted \emph{suṣumṇā\-nāḍī\-sthā}. We are not aware of a reference in another text to impurities in the central channel, however, so we have adopted the reading of the \gamma \ group and \betaTwo, which states that the impurites are situated at the sides of Suṣumṇā, understanding this to refer to the secondary channels of \emph{iḍā} and \emph{piṅgalā}.

%The \emph{ca} at the end of the fourth verse quarter would suggest that two statements are being made in the second half of the verse. The readings of some manuscripts, such as \emph{suṣumṇā susnigdhā} in \epsilonThree\ and \etaOne, appear to be attempts to make sense of the \emph{ca}. However, the meaning of \emph{suṣumṇā susnigdhā} (`\emph{suṣumṇā} becomes well-lubricated') seems implausible. In light of the reading in the \gamma \ group and \betaTwo, it seems likely that \emph{ca} was simply a verse-filler.
%JM: I think we should adopt alpha1 (and others): suṣumṇā susvasthā. It works, and makes sense of ca. But we'd not be reading alpha1's yāsthā, which looks to be a mistake.
\end{philcomm}

@@

%\begin{philcomm}[hp02_007]
%The hemistich beginning with \emph{dhārayitvā} appears to have dropped out of the transmission of the \emph{Haṭharatnāvalī}, thus rendering this passage unintelligible in that work (see Gharote et al. 2002: 125).%JM: necessary?
%\end{philcomm}

@@

\emph{Yukabhavadeva} 7.14

@@

\begin{sources}[hp02_009]
Cf. \emph{Dattātreyayogaśāstra} 61

@@

\begin{philcomm}[hp02_009]
The meaning of \emph{avirodhataḥ} (‘without harm/discomfort’) makes better sense in this context than \emph{anirodhataḥ} (‘without cessation’). One might try to construe \emph{anirodhataḥ} as ‘without stopping the breath,’ but verse 2.7 clearly states that the breath should be held as long as possible (\emph{yathāśakti}). According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested for the parallel hemistich. Furthermore, \emph{avirodhata} is attested, as well as \emph{virodhahīna}, in a passage of the \emph{Haṭhatattvakaumudī} (36.6–9) that appears to have been loosely based on \emph{Haṭhapradīpikā} 2.7–9:

@@

\begin{philcomm}[hp02_010]
%We have translated \emph{yamināṃ} as `yogis' because in this context it could refer to those practising the observances (\emph{yama}) of yoga rather than ascetics more generally. In \emph{Jyotsnā} 2.44, Brahmānanda understands it this way (\emph{yamināṃ yogināṃ}).%JM: It says ascetics, so why not translate thus?
%The reading \emph{vidhinābhyāsaṃ} in the third pāda of various witnesses is unmetrical because it lacks the caesura. (not well attested in the new collation). 
\end{philcomm}

@@

\begin{philcomm}[hp02_012]
The manuscript readings diverge greatly in the second hemistich. In the third verse quarter, all of the manuscripts have \emph{prāṇa} in some form (instead of \emph{deha} in the \emph{Vivekamārtaṇḍa}). \textalpha\ and some of the \textbeta, \textepsilon\ and \texteta\ manuscripts seem to be stating that it is \emph{padmāsana} that rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}). We have adopted a similar reading but with \emph{padmāsane} (\betaTwo, \deltaOne\ and \deltaThree, and \etaTwo) because it makes better sense that the yogi rises up while seated in lotus pose. Another version is seen in \etaTwo\ (and others), which appears to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, the verse is about the external signs that might arise in \emph{prāṇāyāma} rather than internal processes. Such confusion has arisen because the verse was taken from the \emph{Vivekamārtaṇḍa} without its context, which is a classification of different levels of \emph{prāṇāyāma}, so Svātmārāma needed to include \emph{prāṇarodhe} meaning \emph{prāṇāyāma} in order for the different adjectives to have something with which to agree, and he did so despite the infelicity of \emph{prāṇarodhe} crossing the \emph{pāda} break.
\end{philcomm}

@@

\begin{philcomm}[hp02_014]
\end{philcomm}

@@

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śānaiḥ |\\+}
\tl{anyathā hanti yantāraṃ tathā vāyur asevitaḥ ||\\!}
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp02_015]
The second hemistich of this verse has been rewritten to express the same idea (and simile) as that found in \emph{Viveka\-mārtaṇḍa} 101, but the author of the \emph{Viveka\-mārtaṇḍa} expresses it more clearly.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.18}
\begin{translation}[hp02_018]
[The yogi] should exhale correctly, inhale correctly and hold the breath correctly. He thus becomes purified.
\end{translation}

@@

\emph{Yogacintāmaṇi} 92v (attr.~to the \emph{Skandpurāṇa})

@@

%\begin{philcomm}[hp02_018]
%Some of the alpha, epsilon and eta manuscripts have \emph{śuddhiṃ} instead of \emph{siddhiṃ} in the fourth verse quarter but this is likely a later change, or even an error, by someone who was anticipating the subject of the next verse. The reading \emph{siddhiṃ} is attested across all branches of the stemma, as well as the source text and most of the testimonia.

@@

\begin{philcomm}[hp02_021]
Manuscripts of the delta group, as well as \etaTwo\ and \epsilonThree, also have the valid readings of \emph{medaḥśleṣmādināśārthaṃ} and \emph{anyathā} in the first and third verse quarters, respectively. This version of the verse states that one should practise the six therapeutic interventions to remove fat, phlegm and the like, otherwise one should not practise them when the humours are in balance. However, the \emph{pūrvam} (`first'), which is attested in two \textalpha\ manuscripts, as well as manuscripts of the \textbeta\ and \textgamma\ groups, fits the context of these interventions being preliminary practices for \emph{prāṇāyāma}.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.22}
\begin{translation}[hp02_022]
\emph{Dhauti, basti, neti, trāṭaka, naulī} and \emph{kapālabhātī}. These are said to be the six [therapeutic] techniques.%JM: "therapeutic interventions" adds a lot to karmāṇi. I usually translate it with "[cleansing] techniques". We have techniques everywhere else.
%JM: note on alphaone's laulikam? And should we have naulī here: that's what it is everywhere else so this is prob for metre.
\end{translation}

@@

\begin{versinnote}
\tl{haṭhapradīpikāyām-\\+}
\tl{dhautir bastis tathā netis trāṭakaṃ naulikaṃ tathā |\\+}
\tl{kapālabhrāntir etāni ṣaṭkarmāṇi pracakṣate  ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp02_023]
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
{}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
\end{translation}

@@

\begin{philcomm}[hp02_024]
The manuscripts contain many different readings for the fourth \emph{pāda}. \alphaThree\ and \etaOne\ have the term \emph{udgāraṃ}, which rarely occurs in yoga texts. The basic meaning of \emph{udgāra} is the act of discharging something from the mouth, which fits the context of \emph{dhauti} in so far as the cloth swallowed into the stomach is drawn back out through the mouth. Some of the other readings, such as \emph{uditaṃ}, \emph{utthānaṃ}, etc., appear to be mistakes or patches that arose possibly because \emph{udgāra} is not normally neuter but masculine.

Many manuscripts have added verse quarters on the length of the cloth (\emph{hastapañcadaśena tu}) and doing the practice according to the guru’s teachings (\emph{gurūpadiṣṭa\-mārgeṇa}). These additional comments are absent in the \textalpha, \textgamma\ and \textepsilon\ groups. Furthermore, the compound \emph{hastapañcadaśena} does not seem to fit the syntax of the sentence. The other addition, on the guru's teaching, is a cliché that is probably being used here as a verse filler.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, are sure to disappear through the power of the dhauti technique.
\end{translation}

@@

\begin{philcomm}[hp02_025]
The verb \emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “purify”.

@@

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
\end{translation}
% Adopt pakhālaṃ — not done in latest version

@@

\begin{philcomm}[hp02_026]
% MD: kṣālanaṃ or prakṣālaṃ? Adopt pakhālaṃ [MD: done]
The reading \emph{pakhālaṃ} that we have adopted in the fourth \emph{pāda} is found in \betaOmega \ and \alphaOne \ and reflects vernacular usage as found in the old Hindi \emph{Aṣṭāṅgayoga} of Caraṇadāsa (6.71ab). Some other witnesses have the Sanskritised  form \emph{prakṣālaṃ}, which is very rare in Sanskrit sources, while others have the more common \emph{kṣālanaṃ}.

@@

\begin{versinnote}
\tl{dhātvīndrintaḥ karaṇaprabodhaṃ\\+}
\tl{dadāti kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād\\+}
\tl{abhyasyamānaṃ jalavastikarma ||\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.30 heading}
\begin{translation}[hp02_030a]
Now \emph{neti}.
\end{translation}

>>>>

\begin{versinnote}
\tl{atha prāṇāyāmaḥ--\\+}
\tl{athāsane dṛḍhe yogī vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prāṇāyāmān samabhyaset ||\\!}
\end{versinnote}

@@

\begin{versinnote}
\tl{atha nāḍīśuddhiḥ--\\+}
\tl{tathā coktaṃ yogacandrikāyām--\\+}
\tl{athāsane dṛḍhībhūte vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prāṇāyāmān samabhyaset ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp02_001]
%\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc. (not worth commenting on as alpha has dṛḍhe).
The \emph{Jyotsnā} (2.1) has the plural \emph{prāṇāyāmān}, which is supported by some manuscripts of the \textbeta, \textgamma\ and \textzeta\ groups. The plural is possible here as it could refer to the different techniques of retention (\emph{kumbhaka}) taught later in the chapter. This is how Brahmānanda understands it (\emph{prāṇāyāmān vakṣyamāṇān}). The variation between singular and plural recurs through this chapter, and we have followed the \alphaOne\ readings, which make good sense. In this case, since the verse is introducing the topic of breath control, the more general sense of the singular is appropriate.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind is moving, everything moves [and] when it is still, everything is firmly fixed, so the yogi attains motionlessness through restraining the breath.
\end{translation}

@@

\begin{philcomm}[hp02_002]
%The \textalpha\ manuscripts, as well as those of various other important groups, have \emph{sarvaṃ} instead of \emph{cittaṃ} in the first verse quarter. The stemma aside, the former makes better sense as a general statement (i.e., when the wind moves, everything moves, etc.). The reading \emph{citta} only seems to make sense if one understands \emph{sthāṇutvam} as referring to \emph{samādhi} (i.e., `motionlessness [of the mind]'). As far as we know, such a meaning of \emph{sthāṇutvam} is not attested elsewhere. Nonetheless, Brahmānanda glosses \emph{sthāṇutvam} as \emph{kāṣṭhavat} (`like a piece of wood'), implying that it refers to a yogi in \emph{samādhi} as this simile is found elsewhere in contexts of meditative absorption (e.g., \emph{Amanaska} 1.27).
%(JM: but we have sarvam now and cittam is clearly wrong; I think this note is unnecessary — sthāṇu can mean a piece of wood or a post and is a byword for motionlessness (but never means samādhi), like kāṣṭha, and we don't want to put people off reading the notes by including too many unedifying ones.) 
% JB: Since we often read against the stemma, I dont think it's very convincing to simply say we are reading with the stemma in cases where there are two well-attested plausible readings. In this particular case, cittaṃ is a plausible reading that's well attested among the collated mss., so I think we should firstly bring it to the reader's attention and then explain why we think the adopted reading is better on semantic grounds. Perhaps, my attempts to do the latter (by tidying up a previous note) can be improved?
%JM: should we mention that sthāṇu can mean Śiva?
In the second verse quarter, most of the \textalpha, \textbeta\ and \textgamma\ manuscripts have \emph{dṛḍhabandhanaṃ} instead of the reading we have adopted, \emph{niścalaṃ tathā}, which is supported by some \textbeta, \textdelta\ and \texteta\ manuscripts, and also the source text (i.e. the \emph{Vivekamārtaṇḍa}) and testimonia. The reading \emph{dṛḍhabandhanaṃ} appears to be a dittographical error and does not make sense when read with either \emph{sarvaṃ} (\textalpha) or \emph{cittaṃ} (\textgamma). \lb

%MD 2024-8: The most plausible reconstruction of the text, based on our manuscripts, is "cale vāte calet sarvaṃ niścale dṛḍhabandhanam" (When the wind moves, everything would move. When [the wind] is still, [everything] would have firm bond.) Would this work? Yes, with calaṃ ?? JM to rewrite note.

As noted by Brahmānanda, in addition to its primary meaning of immobility, \emph{sthāṇutvam} can also mean the state of being Śiva, for whom Sthāṇu is another name. \lb

@@

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate |\\+}
\tl{maraṇaṃ tasya niṣkrāntau tato vāyuṃ nirodhayet ||\\+}
\tl{\var{72a sthito ] sthiro A 72b jīvitam ] VAGBGPT; jīvanam GLGPk 72c niṣkrāntau ] VTA; niḥkrāṃtaṃ GBGP, niṣkrāntis GLGPk 72d vāyuṃ nirodhayet ] T; vāyunibandhanāt V, vāyuṃ nibandhayet ATvl, vāyuṃ nirundhayet G}\\!}
\end{versinnote}

@@

\begin{versinnote}
\tl{vyāpārād yasya ceṣṭante śārīrāḥ pañca vāyavaḥ ||\\+}
\tl{prāṇāpānādayas te tu bhinnā vṛtter na vastutaḥ |\\+}
\tl{vṛttiṃ leśān nigadato bharadvāja nibodha me ||\\+}
\tl{vṛttiḥ prāṇāyānaṃ nāma yat taj jīvanam ucyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_003]
\emph{Haṭharatnāvalī} 3.80

@@

%We believe that \emph{nirundhayet} was probably the original reading and that it was changed to the more correct form of \emph{nirodhayet} and, in some cases, confused with the final verse quarter of the previous verse and changed to \emph{nibandhanāt} (\alphaOne) or \emph{nibandhayet} (\alphaThree\textbeta).%JM: necessary note?
%JB nirundhayet is not supported by alpha.
%MD: Why is the reading vāyuṃ nibandhayet not possible, when we have vāyunibandhanāt in 2.2d? See also samīraṇam alaṃ baddhvā in 2.10b. The development of the variant readings "nibandhayet -> nirundhayet -> nirodhayet" is much more plausible.
% JB I've commmented on jīvitaṃ/jīvanam because it looks like its unsupported by alpha (alpha one has jītaviṃ). I now see that J5 has jīvitaṃ, so we could eliminate the note if we add this to the collation. But a note might still be helpful here because G4 is damaged at this place, so jīvitaṃ is not so well attested and jīvanaṃ is possible. 
\end{philcomm}

@@

%\begin{philcomm}[hp02_004]
%The reading \emph{kāyasiddhi} in the final verse quarter is supported by manuscripts of the \textalpha, \textbeta, \textzeta\ and \texteta\ groups whereas other manuscripts have \emph{kāryasiddhi}. In the context of physical yoga, \emph{kāyasiddhi} makes better sense, as \emph{kāryasiddhi} can refer more generally to accomplishing anything.
%JM: necessary for the reader? Goes with the stemma, makes better sense, variant not particularly remarkable. But maybe I have become too parsimony-partial.
% JB. In this case I agree!
%\end{philcomm}

@@

%\begin{philcomm}[hp02_005]
%We have understood the term \emph{nāḍīcakra} as the network of channels that originate from the bulb in the lower abdomen. This interpretation is consistent with \emph{Sārdhatriśatikālottara} 10.1–3:%JM there are many many possible refs for this, do we need one? And some systems have them coming from the navel, and some from the heart (Shaman's new article).

@@

%\begin{philcomm}[hp02_006]
%There is a problem in the third verse quarter. Among the divergent readings, the delta manuscripts convey the idea that the impurities are in the central channel  (\emph{suṣumṇāntarasthā}), which makes a defective \emph{ra-vipulā} (there is no caesura after the fourth syllable). This is similar in sense to V2(siglum), which has the reading \emph{suṣumṇā\-madhya\-sthā}. This meaning was accepted by Brahmānanda, who adopted \emph{suṣumṇā\-nāḍī\-sthā}. We are not aware of a reference in another text to impurities in the central channel, however, so we have adopted the reading of the \gamma \ group and \betaTwo, which states that the impurites are situated at the sides of Suṣumṇā, understanding this to refer to the secondary channels of \emph{iḍā} and \emph{piṅgalā}.

%The \emph{ca} at the end of the fourth verse quarter would suggest that two statements are being made in the second half of the verse. The readings of some manuscripts, such as \emph{suṣumṇā susnigdhā} in \zetaThree\ and \etaOne, appear to be attempts to make sense of the \emph{ca}. However, the meaning of \emph{suṣumṇā susnigdhā} (`\emph{suṣumṇā} becomes well-lubricated') seems implausible. In light of the reading in the \gamma \ group and \betaTwo, it seems likely that \emph{ca} was simply a verse-filler.
%JM: I think we should adopt alpha1 (and others): suṣumṇā susvasthā. It works, and makes sense of ca. But we'd not be reading alpha1's yāsthā, which looks to be a mistake.
%\end{philcomm}

@@

%\begin{philcomm}[hp02_007]
%The line beginning with \emph{dhārayitvā} appears to have dropped out of the transmission of the \emph{Haṭharatnāvalī}, thus rendering this passage unintelligible in that work (see Gharote et al. 2002: 125).%JM: necessary?
%\end{philcomm}

@@

\emph{Yuktabhavadeva} 7.14

@@

\begin{sources}[hp02_009]
Cf.~\emph{Dattātreyayogaśāstra} 61

@@

\begin{philcomm}[hp02_009]
The meaning of \emph{avirodhataḥ} (‘without harm/discomfort’) makes better sense in this context than \emph{anirodhataḥ} (‘without cessation’). One might try to construe \emph{anirodhataḥ} as ‘without stopping the breath,’ but verse 2.7 clearly states that the breath should be held as long as possible (\emph{yathāśakti}). According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested for the parallel line. Furthermore, \emph{avirodhata} is attested, as well as \emph{virodhahīna}, in a passage of the \emph{Haṭhatattvakaumudī} (36.6–9) that appears to have been loosely based on \emph{Haṭhapradīpikā} 2.7–9:

@@

%\begin{philcomm}[hp02_010]
%We have translated \emph{yamināṃ} as `yogis' because in this context it could refer to those practising the observances (\emph{yama}) of yoga rather than ascetics more generally. In \emph{Jyotsnā} 2.44, Brahmānanda understands it this way (\emph{yamināṃ yogināṃ}).%JM: It says ascetics, so why not translate thus?
%The reading \emph{vidhinābhyāsaṃ} in the third pāda of various witnesses is unmetrical because it lacks the caesura. (not well attested in the new collation). 
%\end{philcomm}

@@

\begin{philcomm}[hp02_012]
The manuscript readings diverge greatly in the second line. In the third verse quarter, all of the manuscripts have \emph{prāṇa} in some form (instead of \emph{deha} in the \emph{Vivekamārtaṇḍa}). \textalpha\ and some of the \textbeta, \textzeta\ and \texteta\ manuscripts seem to be stating that it is \emph{padmāsana} that rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}). We have adopted a similar reading but with \emph{padmāsane} (\betaTwo, \deltaOne, and \etaTwo) because it makes better sense that the yogi rises up while seated in lotus pose. Another version is seen in \etaTwo, which appears to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, the verse is about the external signs that might arise in \emph{prāṇāyāma} rather than internal processes. Such confusion has arisen because the verse was taken from the \emph{Vivekamārtaṇḍa} without its context, which is a classification of different levels of \emph{prāṇāyāma}, so Svātmārāma needed to include \emph{prāṇarodhe} meaning \emph{prāṇāyāma} in order for the different adjectives to have something with which to agree, and he did so despite the infelicity of \emph{prāṇarodhe} crossing the \emph{pāda} break.
\end{philcomm}

@@

% \begin{philcomm}[hp02_014]
% \end{philcomm}

@@

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ |\\+}
\tl{anyathā hanti yantāraṃ tathā vāyur asevitaḥ ||\\!}
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp02_015]
The second line of this verse has been rewritten to express the same idea (and simile) as that found in \emph{Viveka\-mārtaṇḍa} 101, but the author of the \emph{Viveka\-mārtaṇḍa} expresses it more clearly.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.18}
\begin{translation}[hp02_018]
[The yogi] should exhale correctly, inhale correctly and hold the breath correctly. [His channels] thus become purified.
\end{translation}

@@

\emph{Yogacintāmaṇi} 92v (attr.~to the \emph{Skandapurāṇa})

@@

%\begin{philcomm}[hp02_018]
%Some of the alpha, zeta and eta manuscripts have \emph{śuddhiṃ} instead of \emph{siddhiṃ} in the fourth verse quarter but this is likely a later change, or even an error, by someone who was anticipating the subject of the next verse. The reading \emph{siddhiṃ} is attested across all branches of the stemma, as well as the source text and most of the testimonia.

@@

\begin{philcomm}[hp02_021]
Manuscripts of the \textdelta\ group, as well as \etaTwo\ and \zetaThree, also have the valid readings of \emph{medaḥśleṣmanivṛtyarthaṃ} (\deltaOne\deltaTwo\zetaThree)/\emph{medaḥśleṣmādināśārthaṃ} (\etaTwo) and \emph{anyathā} in the first and third verse quarters, respectively. This version of the verse states that one should practise the six therapeutic interventions to remove fat, phlegm and the like, otherwise one should not practise them when the humours are in balance. However, the \emph{pūrvam} (`first'), which is attested in the \textalpha, \textbeta\ and \textgamma\ groups, fits the context of these interventions being preliminary practices for \emph{prāṇāyāma}.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.22}
\begin{translation}[hp02_022]
\emph{Dhauti, basti, nīti, trāṭaka, naulī} and \emph{kapālabhātī}. These are said to be the six [therapeutic] techniques.%JM: "therapeutic interventions" adds a lot to karmāṇi. I usually translate it with "[cleansing] techniques". We have techniques everywhere else.
%JM: note on alphaone's laulikam? And should we have naulī here: that's what it is everywhere else so this is prob for metre.
\end{translation}

@@

\begin{versinnote}
\tl{haṭhapradīpikāyām-\\+}
\tl{dhautir bastis tathā netis trāṭakaṃ naulikaṃ tathā |\\+}
\tl{kapālabhātir etāni ṣaṭkarmāṇi pracakṣate  ||\\!}     %  kapālabhrāntir   corrected
\end{versinnote}

@@

% \begin{philcomm}[hp02_023]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
{[}The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
\end{translation}

@@

\begin{philcomm}[hp02_024]
The manuscripts contain many different readings for the fourth \emph{pāda}. \alphaThree\ and \etaOne\ have the term \emph{udgāraṃ}, which rarely occurs in yoga texts. The basic meaning of \emph{udgāra} is the act of discharging something from the mouth, which fits the context of \emph{dhauti} in so far as the cloth swallowed into the stomach is drawn back out through the mouth. Some of the other readings, such as \emph{uditaṃ}, \emph{utthānaṃ}, etc., appear to be mistakes or patches that arose possibly because \emph{udgāra} is not normally neuter but masculine. \lb
% MD: Why don't we adopt the reading ukhāla? ukhāla seems to mean "vomiting" in Hindi, Marathi, Nepali etc. See also https://www.wisdomlib.org/definition/ukhala-pakhala. We have accepted pakhāla in 2.26. Apparently they belong together. ??JM to write note

Many manuscripts have added verse quarters on the length of the cloth (\emph{hastapañcadaśena tu}) and doing the practice according to the guru’s teachings (\emph{gurūpadiṣṭa\-mārgeṇa}). These additional comments are absent in the \textalpha, \textgamma\ and \textzeta\ groups. Furthermore, the compound \emph{hastapañcadaśena} does not seem to fit the syntax of the sentence. The other addition, on the guru's teaching, is a cliché that is probably being used here as a verse filler.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, are sure to flee (\textit{dhāvanti}) through the power of the dhauti technique.
\end{translation}

@@

\begin{philcomm}[hp02_025]
The verb \emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “purify”. \lb

@@

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
\end{translation}

@@

\begin{philcomm}[hp02_026]
The reading \emph{pakhālaṃ} that we have adopted in the fourth \emph{pāda} is found in \alphaTwo\ and \betaOmega\ and is close to the reading of \alphaOne. %  not exactly: pākhālaṃ α1, paṣālaṃ α2
It reflects vernacular usage as found in the old Hindi \emph{Aṣṭāṅgayoga} of Caraṇadāsa (6.71ab). Some other witnesses have the Sanskritised form \emph{prakṣālaṃ}, which is very rare in Sanskrit sources, while others have the more common \emph{kṣālanaṃ}.

@@

\begin{versinnote}
\tl{dhātvīndriyāntaḥkaraṇaprabodhaṃ\\+}       %  dhātvīndrintaḥ corrected JH
\tl{dadāti kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād\\+}
\tl{abhyasyamānaṃ jalavastikarma ||\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.30 heading}
\begin{translation}[hp02_030a]
Now \emph{nīti}.
\end{translation}
